---
title: "aDNA"
author: "Jennifer Blanc"
date: "2024-07-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(data.table)
library(tidyverse)
library(ggpubr)
library(latex2exp)
```


# Construct Test and GWAS panels 

## Explore data

```{r}
df <- fread("../data/adnaSel/raw/all_populations_mles.txt")
head(df)
```

### Overlap statistics

Load data and format 
```{r}
dfOverlap <- fread("../endpoints/adnaSel/OverlapStats.txt") %>% separate(gwas, into = c("tmp", "gwas"), sep = "e")
dfOverlap$gwas <- factor(dfOverlap$gwas, levels = c( "5", "10", "50", "200"))
```

Plot Z by GWAS
```{r}
pl <- dfOverlap %>% filter(L == "pruneall")%>% ggplot(aes(x = gwas, y = Z)) + geom_point() + xlab("") + ylab("") + theme_bw(base_size = 14)  + guides(fill="none") + theme(plot.title = element_text(hjust = 0.5), axis.text.x = element_text(size = 10), strip.text.x = element_text(size = 8,))+facet_wrap(~contrasts, nrow = 1)+ scale_y_log10(limits = c(1e-6, 1e-4)) 
pl
ggsave("~/Desktop/tmp.png", pl, height = 2, width = 12)
```



**GWAS type vs proportion of variance explained** 
```{r}
EN <- dfOverlap %>% filter(contrasts == "EuropeanNeolithic") %>% ggplot(aes(x = gwas, y = Z, fill = gwas_type2)) + geom_bar(stat = "identity") + xlab("GWAS Type") + ylab("Proportion of variance explained (Z)") + theme_bw(base_size = 14) + scale_fill_manual(values = c("purple4", "lightseagreen"))  + guides(fill="none") + ggtitle("Neolithic") + theme(plot.title = element_text(hjust = 0.5)) + ylim(0, 7e-5)

BA <- dfOverlap %>% filter(contrasts == "BronzeAge") %>% ggplot(aes(x = gwas, y = Z, fill = gwas_type2)) + geom_bar(stat = "identity") + xlab("GWAS Type") + ylab("Proportion of variance explained (Z)") + theme_bw(base_size = 14) + scale_fill_manual(values = c("purple4", "lightseagreen"))  + guides(fill="none") + ggtitle("Bronze Age") + theme(plot.title = element_text(hjust = 0.5))+ ylim(0, 7e-5)

H <- dfOverlap %>% filter(contrasts == "Historical") %>% ggplot(aes(x = gwas, y = Z, fill = gwas_type2)) + geom_bar(stat = "identity") + xlab("GWAS Type") + ylab("Proportion of variance explained (Z)") + theme_bw(base_size = 14) + scale_fill_manual(values = c("purple4", "lightseagreen"))  + guides(fill="none") + ggtitle("Historical") + theme(plot.title = element_text(hjust = 0.5))+ ylim(0, 7e-5)

# Print plots
p <- ggarrange(EN, BA, H, nrow = 1)
p
```
The Bronze age contrasts explain a larger proportion of variance in both ALL and WBS compared to the other two sets of contrasts. Bronze age is modeled as a mixture of Steppe and European Neolithic and has the largest sample size in their analysis. Compared to using q1-q4 as quartiles, using distances generates a smooth gradient between WBS and ALL. 

**GWAS type vs error** 
```{r}
EN <- dfOverlap %>% filter(contrasts == "EuropeanNeolithic") %>% ggplot(aes(x = gwas, y = error, fill = gwas_type2)) + geom_bar(stat = "identity") + xlab("GWAS Type") + ylab("Error") + theme_bw(base_size = 14) + scale_fill_manual(values = c("purple4", "lightseagreen"))  + guides(fill="none") + ggtitle("Neolithic") + theme(plot.title = element_text(hjust = 0.5)) + ylim(0,1) + geom_hline(yintercept = 0, color = "red", linetype = 2, linewidth = 1.2)

BA <- dfOverlap %>% filter(contrasts == "BronzeAge") %>% ggplot(aes(x = gwas, y = error, fill = gwas_type2)) + geom_bar(stat = "identity") + xlab("GWAS Type") + ylab("Error") + theme_bw(base_size = 14) + scale_fill_manual(values = c("purple4", "lightseagreen"))  + guides(fill="none") + ggtitle("BronzeAge") + theme(plot.title = element_text(hjust = 0.5)) + ylim(0,1) + geom_hline(yintercept = 0, color = "red", linetype = 2, linewidth = 1.2)

H <- dfOverlap %>% filter(contrasts == "Historical") %>% ggplot(aes(x = gwas, y = error, fill = gwas_type2)) + geom_bar(stat = "identity") + xlab("GWAS Type") + ylab("Error") + theme_bw(base_size = 14) + scale_fill_manual(values = c("purple4", "lightseagreen"))  + guides(fill="none") + ggtitle("Historical") + theme(plot.title = element_text(hjust = 0.5)) + ylim(0,1) + geom_hline(yintercept = 0, color = "red", linetype = 2, linewidth = 1.2)


# Print plots
p <- ggarrange(EN, BA, H, nrow = 1)
p
```
Similar to other analyses, there is lower error when more ancestries are included. Unlike other analyses, the error is quite a bit higher for both ALL and WBS.  

**Proportion variance explained vs error**   
```{r}
EN <- dfOverlap %>% filter(contrasts == "EuropeanNeolithic") %>% ggplot(aes(x = Z, y = error, color = gwas)) + geom_point(size = 3) + xlab("Z") + ylab("Error") + theme_bw(base_size = 12)  + guides(fill="none") + ggtitle("Neolithic") + theme(plot.title = element_text(hjust = 0.5), legend.position = "bottom") + ylim(0,1) + labs(color = "GWAS Type") + geom_hline(yintercept = 0, color = "red", linetype = 2, linewidth = 1.2)

BA <- dfOverlap %>% filter(contrasts == "BronzeAge") %>% ggplot(aes(x = Z, y = error, color = gwas)) + geom_point(size = 3) + xlab("Z") + ylab("Error") + theme_bw(base_size = 12)  + guides(fill="none") + ggtitle("Bronze Age") + theme(plot.title = element_text(hjust = 0.5), legend.position = "bottom") + ylim(0,1) + labs(color = "GWAS Type") + geom_hline(yintercept = 0, color = "red", linetype = 2, linewidth = 1.2)

H <- dfOverlap %>% filter(contrasts == "Historical") %>% ggplot(aes(x = Z, y = error, color = gwas)) + geom_point(size = 3) + xlab("Z") + ylab("Error") + theme_bw(base_size = 12)  + guides(fill="none") + ggtitle("Historical") + theme(plot.title = element_text(hjust = 0.5), legend.position = "bottom") + ylim(0,1) + labs(color = "GWAS Type") + geom_hline(yintercept = 0, color = "red", linetype = 2, linewidth = 1.2)

# Print plots
p <- ggarrange(EN, BA, H, nrow = 1, common.legend = TRUE)
p
```

Combine into one plot 
```{r}
dfOverlap %>% ggplot(aes(x = Z, y = error, color = gwas, shape = contrasts)) + geom_point(size = 3) + xlab("Z") + ylab("Error") + theme_bw(base_size = 12)  + guides(fill="none")+ theme(plot.title = element_text(hjust = 0.5), legend.position = "bottom") + ylim(0,1) + labs(color = "GWAS Type") + geom_hline(yintercept = 0, color = "red", linetype = 2, linewidth = 1.2)
```
In general, the larger the Z the smaller the error but this is the first example that it doesn't strictly hold. 

**GWAS type vs D**   
```{r}
dfOverlap$d_lc <- dfOverlap$D - (2*sqrt(dfOverlap$varD)) 
dfOverlap$d_uc <- dfOverlap$D + (2 *sqrt(dfOverlap$varD))

EN <- dfOverlap %>% filter(contrasts == "EuropeanNeolithic") %>% ggplot(aes(x = gwas, y = D, color = gwas_type2)) + geom_point(size = 3) + xlab("GWAS Type") + ylab("D") + theme_bw(base_size = 14) + scale_color_manual(values = c("purple4", "lightseagreen"))  + guides(color="none") + ggtitle("Neolithic") + theme(plot.title = element_text(hjust = 0.5)) + geom_hline(yintercept = 0, color = "red", linetype = 2, linewidth = 1.2) + geom_errorbar(aes(ymin = d_lc, ymax = d_uc)) + ylim(0, 34000000)

BA <- dfOverlap %>% filter(contrasts == "BronzeAge") %>% ggplot(aes(x = gwas, y = D, color = gwas_type2)) + geom_point(size = 3) + xlab("GWAS Type") + ylab("D") + theme_bw(base_size = 14) + scale_color_manual(values = c("purple4", "lightseagreen"))  + guides(color="none") + ggtitle("Bronze Age") + theme(plot.title = element_text(hjust = 0.5)) + geom_hline(yintercept = 0, color = "red", linetype = 2, linewidth = 1.2) + geom_errorbar(aes(ymin = d_lc, ymax = d_uc)) + ylim(0, 34000000)

H <- dfOverlap %>% filter(contrasts == "Historical") %>% ggplot(aes(x = gwas, y = D, color = gwas_type2)) + geom_point(size = 3) + xlab("GWAS Type") + ylab("D") + theme_bw(base_size = 14) + scale_color_manual(values = c("purple4", "lightseagreen"))  + guides(color="none") + ggtitle("Historical") + theme(plot.title = element_text(hjust = 0.5)) + geom_hline(yintercept = 0, color = "red", linetype = 2, linewidth = 1.2) + geom_errorbar(aes(ymin = d_lc, ymax = d_uc)) + ylim(0, 34000000)

# Print plots
p <- ggarrange(EN, BA, H, nrow = 1, common.legend = TRUE)
p
```

This is not a great plot because the GWAS panels are different sizes (q1-q4 are all the same) so it's not a fair comparison but plotting the expectation just looks like its on the 0 line because $M-1$ is small compared to $D$. All of the p-value for D are signficant though there is significant variability in the order of magnitude. 


```{r}
dfOverlap %>% select(D, pvalD, contrasts, gwas)
```


### PC statistics

Load and format results
```{r}
dfPC <- fread("../endpoints/adnaSel/PCStats.txt")
dfPC <- dfPC %>% mutate(gwas_sample = case_when(gwas %in% c("q1", "q2", "q3", "q4") ~ "distance",
         gwas %in% c("ALL", "WBS") ~ "full"))
dfPC$gwas <- factor(dfPC$gwas, levels = c("ALL", "WBS", "q1", "q2", "q3", "q4"))
```

**Squared correlation between $PC_i$ between $\hat{F}_{Gr}$**
```{r}
# EN
EN <- dfPC %>% filter(gwas_type == "even" & test_type == "odd" & contrasts == "EuropeanNeolithic")%>% ggplot(aes(x = PC, y = B2)) + geom_bar(stat = "identity", fill = "darkgreen", color = "darkgreen") + facet_grid(~gwas) + theme_bw(base_size = 12) + ylab(TeX("$B_{i}^2$")) + ggtitle("Neolithic") + theme(plot.title = element_text(hjust = 0.5)) + ylim(0,0.7)

# BA
BA <- dfPC %>% filter(gwas_type == "even" & test_type == "odd" & contrasts == "BronzeAge")%>% ggplot(aes(x = PC, y = B2)) + geom_bar(stat = "identity",  fill = "darkgreen", color = "darkgreen") + facet_grid(~gwas) + theme_bw(base_size = 12) + ylab(TeX("$B_{i}^2$")) + ggtitle("Bronze Age") + theme(plot.title = element_text(hjust = 0.5))+ ylim(0,0.7)

# H
H <- dfPC %>% filter(gwas_type == "even" & test_type == "odd" & contrasts == "Historical")%>% ggplot(aes(x = PC, y = B2)) + geom_bar(stat = "identity",  fill = "darkgreen", color = "darkgreen") + facet_grid(~gwas) + theme_bw(base_size = 12) + ylab(TeX("$B_{i}^2$")) + ggtitle("Historical") + theme(plot.title = element_text(hjust = 0.5))+ ylim(0,0.7)

# Print plots
p <- ggarrange(EN, BA, H, nrow = 3, common.legend = TRUE)
p
```

Limit to "ALL"
```{r}
# EN
EN <- dfPC %>% filter(gwas_type == "even" & test_type == "odd" & contrasts == "EuropeanNeolithic" & gwas == "ALL")%>% ggplot(aes(x = PC, y = B2)) + geom_bar(stat = "identity",  fill = "darkgreen", color = "darkgreen") + facet_grid(~gwas) + theme_bw(base_size = 12) + ylab(TeX("$B_{i}^2$")) + ggtitle("Neolithic") + theme(plot.title = element_text(hjust = 0.5)) + ylim(0,0.7)

# BA
BA <- dfPC %>% filter(gwas_type == "even" & test_type == "odd" & contrasts == "BronzeAge"& gwas == "ALL")%>% ggplot(aes(x = PC, y = B2)) + geom_bar(stat = "identity",  fill = "darkgreen", color = "darkgreen") + facet_grid(~gwas) + theme_bw(base_size = 12) + ylab(TeX("$B_{i}^2$")) + ggtitle("Bronze Age") + theme(plot.title = element_text(hjust = 0.5)) + ylim(0,0.7)

# H
H <- dfPC %>% filter(gwas_type == "even" & test_type == "odd" & contrasts == "Historical"& gwas == "ALL")%>% ggplot(aes(x = PC, y = B2)) + geom_bar(stat = "identity", fill = "darkgreen", color = "darkgreen") + facet_grid(~gwas) + theme_bw(base_size = 12) + ylab(TeX("$B_{i}^2$")) + ggtitle("Historical") + theme(plot.title = element_text(hjust = 0.5))+ ylim(0,0.7)

# Print plots
p <- ggarrange(EN, BA, H, nrow = 1, common.legend = TRUE)
p
```


Limit to "WBS"
```{r}
# EN
EN <- dfPC %>% filter(gwas_type == "even" & test_type == "odd" & contrasts == "EuropeanNeolithic" & gwas == "WBS")%>% ggplot(aes(x = PC, y = B2)) + geom_bar(stat = "identity",  fill = "darkgreen", color = "darkgreen") + facet_grid(~gwas) + theme_bw(base_size = 12) + ylab(TeX("$B_{i}^2$")) + ggtitle("Neolithic") + theme(plot.title = element_text(hjust = 0.5)) + ylim(0,0.03)

# BA
BA <- dfPC %>% filter(gwas_type == "even" & test_type == "odd" & contrasts == "BronzeAge"& gwas == "WBS")%>% ggplot(aes(x = PC, y = B2)) + geom_bar(stat = "identity", fill = "darkgreen", color = "darkgreen") + facet_grid(~gwas) + theme_bw(base_size = 12) + ylab(TeX("$B_{i}^2$")) + ggtitle("Bronze Age") + theme(plot.title = element_text(hjust = 0.5)) + ylim(0,0.03)

# H
H <- dfPC %>% filter(gwas_type == "even" & test_type == "odd" & contrasts == "Historical"& gwas == "WBS")%>% ggplot(aes(x = PC, y = B2)) + geom_bar(stat = "identity", fill = "darkgreen", color = "darkgreen") + facet_grid(~gwas) + theme_bw(base_size = 12) + ylab(TeX("$B_{i}^2$")) + ggtitle("Historical") + theme(plot.title = element_text(hjust = 0.5))+ ylim(0,0.03)

# Print plots
p <- ggarrange(EN, BA, H, nrow = 1, common.legend = TRUE)
p
```


All three sets of contrasts are quite correlated with the first PC in ALL. In the WBS, it seems the bronze age contrasts are correlated with the first PC. Its hard to think about what these mean when the contrasts are a departure from an expectation and not directly an allele frequency contrast.   

**Cumulative variance explained** 
```{r}
# EN
dfFilter <- dfOverlap %>% filter(contrasts == "EuropeanNeolithic") %>% mutate(gwas_sample = gwas)
EN <- dfPC %>% filter(gwas_type == "even" & test_type == "odd" & contrasts == "EuropeanNeolithic") %>% ggplot(aes(x = PC, y = r2)) + geom_bar(stat = "identity", fill = "darkgreen") + facet_grid(~gwas) + theme_bw(base_size = 12) + ylab(TeX("$R_{K}^2$")) + geom_hline(data = dfFilter, aes(yintercept = signal), color = "red", linetype = 2) + ggtitle("Neolithic") + theme(plot.title = element_text(hjust = 0.5)) + ylim(0,1)

# BA
dfFilter <- dfOverlap %>% filter(contrasts == "BronzeAge") %>% mutate(gwas_sample = gwas)
BA <- dfPC %>% filter(gwas_type == "even" & test_type == "odd" & contrasts == "BronzeAge") %>% ggplot(aes(x = PC, y = r2)) + geom_bar(stat = "identity", fill = "darkgreen") + facet_grid(~gwas) + theme_bw(base_size = 12) + ylab(TeX("$R_{K}^2$")) + geom_hline(data = dfFilter, aes(yintercept = signal), color = "red", linetype = 2) + ggtitle("Bonze Age") + theme(plot.title = element_text(hjust = 0.5)) + ylim(0,1)

# Historical
dfFilter <- dfOverlap %>% filter(contrasts == "Historical") %>% mutate(gwas_sample = gwas)
H <- dfPC %>% filter(gwas_type == "even" & test_type == "odd" & contrasts == "Historical") %>% ggplot(aes(x = PC, y = r2)) + geom_bar(stat = "identity", fill = "darkgreen") + facet_grid(~gwas) + theme_bw(base_size = 12) + ylab(TeX("$R_{K}^2$")) + geom_hline(data = dfFilter, aes(yintercept = signal), color = "red", linetype = 2) + ggtitle("Historical") + theme(plot.title = element_text(hjust = 0.5))+ ylim(0,1)

# Print plots
p <- ggarrange(EN, BA, H , nrow = 3, common.legend = TRUE)
p
```

Limit to ALL
```{r}
# EN
dfFilter <- dfOverlap %>% filter(contrasts == "EuropeanNeolithic"& gwas == "ALL") %>% mutate(gwas_sample = gwas)
EN <- dfPC %>% filter(gwas_type == "even" & test_type == "odd" & contrasts == "EuropeanNeolithic"& gwas == "ALL") %>% ggplot(aes(x = PC, y = r2)) + geom_bar(stat = "identity", fill = "darkgreen") + facet_grid(~gwas) + theme_bw(base_size = 12) + ylab(TeX("$R_{K}^2$")) + geom_hline(data = dfFilter, aes(yintercept = signal), color = "red", linetype = 2) + ggtitle("Neolithic") + theme(plot.title = element_text(hjust = 0.5)) + ylim(0,1)

# BA
dfFilter <- dfOverlap %>% filter(contrasts == "BronzeAge"& gwas == "ALL") %>% mutate(gwas_sample = gwas)
BA <- dfPC %>% filter(gwas_type == "even" & test_type == "odd" & contrasts == "BronzeAge" & gwas == "ALL") %>% ggplot(aes(x = PC, y = r2)) + geom_bar(stat = "identity", fill = "darkgreen") + facet_grid(~gwas) + theme_bw(base_size = 12) + ylab(TeX("$R_{K}^2$")) + geom_hline(data = dfFilter, aes(yintercept = signal), color = "red", linetype = 2) + ggtitle("Bonze Age") + theme(plot.title = element_text(hjust = 0.5))+ ylim(0,1)

# Historical
dfFilter <- dfOverlap %>% filter(contrasts == "Historical"& gwas == "ALL") %>% mutate(gwas_sample = gwas)
H <- dfPC %>% filter(gwas_type == "even" & test_type == "odd" & contrasts == "Historical"& gwas == "ALL") %>% ggplot(aes(x = PC, y = r2)) + geom_bar(stat = "identity", fill = "darkgreen") + facet_grid(~gwas) + theme_bw(base_size = 12) + ylab(TeX("$R_{K}^2$")) + geom_hline(data = dfFilter, aes(yintercept = signal), color = "red", linetype = 2) + ggtitle("Historical") + theme(plot.title = element_text(hjust = 0.5))+ ylim(0,1)

# Print plots
p <- ggarrange(EN, BA, H , nrow = 1, common.legend = TRUE)
p
```


Limit to WBS
```{r}
# EN
dfFilter <- dfOverlap %>% filter(contrasts == "EuropeanNeolithic"& gwas == "WBS") %>% mutate(gwas_sample = gwas)
EN <- dfPC %>% filter(gwas_type == "even" & test_type == "odd" & contrasts == "EuropeanNeolithic"& gwas == "WBS") %>% ggplot(aes(x = PC, y = r2)) + geom_bar(stat = "identity", fill = "darkgreen") + facet_grid(~gwas) + theme_bw(base_size = 12) + ylab(TeX("$R_{K}^2$")) + geom_hline(data = dfFilter, aes(yintercept = signal), color = "red", linetype = 2) + ggtitle("Neolithic") + theme(plot.title = element_text(hjust = 0.5)) + ylim(0,1)

# BA
dfFilter <- dfOverlap %>% filter(contrasts == "BronzeAge"& gwas == "WBS") %>% mutate(gwas_sample = gwas)
BA <- dfPC %>% filter(gwas_type == "even" & test_type == "odd" & contrasts == "BronzeAge" & gwas == "WBS") %>% ggplot(aes(x = PC, y = r2)) + geom_bar(stat = "identity", fill = "darkgreen") + facet_grid(~gwas) + theme_bw(base_size = 12) + ylab(TeX("$R_{K}^2$")) + geom_hline(data = dfFilter, aes(yintercept = signal), color = "red", linetype = 2) + ggtitle("Bonze Age") + theme(plot.title = element_text(hjust = 0.5))+ ylim(0,1)

# Historical
dfFilter <- dfOverlap %>% filter(contrasts == "Historical"& gwas == "WBS") %>% mutate(gwas_sample = gwas)
H <- dfPC %>% filter(gwas_type == "even" & test_type == "odd" & contrasts == "Historical"& gwas == "WBS") %>% ggplot(aes(x = PC, y = r2)) + geom_bar(stat = "identity", fill = "darkgreen") + facet_grid(~gwas) + theme_bw(base_size = 12) + ylab(TeX("$R_{K}^2$")) + geom_hline(data = dfFilter, aes(yintercept = signal), color = "red", linetype = 2) + ggtitle("Historical") + theme(plot.title = element_text(hjust = 0.5))+ ylim(0,1)

# Print plots
p <- ggarrange(EN, BA, H , nrow = 1, common.legend = TRUE)
p
```
