---
title: "SDS"
author: "Jennifer Blanc"
date: "2024-07-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(data.table)
library(tidyverse)
```

## Initial look at both SDS data sets

Read in both files
```{r, warning=FALSE, message=FALSE}
UK <- fread("../data/SDS/raw/SDS_UK10K_n3195_release_Sep_19_2016.tab.gz")
Han <- fread("../data/SDS/raw/auto_SDS_norm_sort.txt.gz")
```

Look at formatting
```{r}
head(UK)
```

```{r}
head(Han)
```

Convert Han coordinates to hg19
```{r}
library(rtracklayer)
library(GenomicRanges)

#chain_url <- "http://hgdownload.cse.ucsc.edu/goldenPath/hg38/liftOver/hg38ToHg19.over.chain.gz"
#chain_file <- "hg38ToHg19.over.chain.gz"
#download.file(chain_url, chain_file)

# Download the chain file
chain_file <- "../data/liftover/hg38ToHg19.over.chain.gz"
#download.file(chain_url, chain_file, mode = "wb")

# Uncompress the chain file
chain_file_unzipped <- sub(".gz$", "", chain_file)
#R.utils::gunzip(chain_file, chain_file_unzipped, remove = FALSE)

# Load the chain file
chain <- import.chain(chain_file_unzipped)

# Create a GRanges object with your SNP coordinates
Han$chr <- paste0("chr", Han$chr)
gr <- GRanges(seqnames = Han$chr, ranges = IRanges(start = Han$pos, end = Han$pos))

# Perform the liftOver
lifted_snps <- liftOver(gr, chain)

# Extract the results
converted_snps <- as.data.frame(unlist(lifted_snps))
converted_snps

# Filter out empty liftOver results
non_empty_indices <- elementNROWS(lifted_snps) > 0
filtered_gr <- gr[non_empty_indices]
filtered_lifted_snps <- lifted_snps[non_empty_indices]

# Convert the filtered GRanges and lifted results to data frames
original_coords <- as.data.frame(filtered_gr)
lifted_coords <- as.data.frame(unlist(filtered_lifted_snps))

# Combine the data frames
combined_coords <- data.frame(
  original_chr = original_coords$seqnames,
  original_start = original_coords$start,
  original_end = original_coords$end,
  lifted_chr = lifted_coords$seqnames,
  lifted_start = lifted_coords$start,
  lifted_end = lifted_coords$end
)

# Print the combined coordinates
print(combined_coords)
```
Join lifted over coordinates with SDS values 
```{r}
dfHan <- inner_join(Han, combined_coords, by = c("chr" = "original_chr", "pos" = "original_start"))

#fwrite(dfHan,"../data/SDS/raw/Han_SDS_liftedover.txt.gz", row.names = F, col.names = T, quote = F, sep = "\t")
```

## Sample Individuals for GWAS panel 

Plot PC1 vs PC2 colored by self-identified ethnic background
```{r, warning=FALSE}
# Load data and combine
dfPCA <- fread("../endpoints/ukbb/whole_biobank.eigenvec.gz")
dfEB <- fread("../endpoints/ukbb/EthnicBackground_21000.txt")
df <- inner_join(dfPCA, dfEB)

# Use continental labels 
df <- df %>% mutate(continental = case_when(EthnicBackground_21000 %in% c(4, 4001, 4002, 4003) ~ "Africa", EthnicBackground_21000 %in% c(1, 1001, 1002, 1003) ~ "Europe", EthnicBackground_21000 %in% c(2, 2001, 2002, 2003, 2004) ~ "Mixed", EthnicBackground_21000 %in% c(3, 3001, 3002, 3003, 3004) ~ "Asia"))
```

```{r}
# Make PCA plot color based on self identified Ethnic background
ggplot(data = df, aes(x = PC1, y = PC2, color = continental)) + geom_point() + theme_classic(base_size = 14) + scale_color_manual(values = c("darkred", "navy", "goldenrod4", "gray50"), na.value = "gray70")
```
Use predefined sets 
```{r}
df <- df %>% mutate(POP = continental)

# Use all genotyped individuals 
dfALL <- df %>% select("#FID", "IID", "POP")
fwrite(dfALL,"../data/SDS/ids/gwas_ids/ALL.txt", row.names = F, col.names = T, quote = F, sep = "\t")

# Use only WBS 
WBS <- fread("../data/ukbb/WBS.ids")
dfWBS <- inner_join(df, WBS) %>% select("#FID", "IID", "POP")
fwrite(dfWBS,"../data/SDS/ids/gwas_ids/WBS.txt", row.names = F, col.names = T, quote = F, sep = "\t")
```

Use distance from WBS
```{r}
# Get average of WBS
dfWBS <- inner_join(df, WBS)
avgPC1 <- mean(dfWBS$PC1)
avgPC2 <- mean(dfWBS$PC2)

# Calculate from WBS centroid 
avgPC <- c(avgPC1, avgPC2)
df$distance <- sqrt(rowSums((df[, 3:4] - avgPC)^2))

# Plot distance
ggplot(data = df, aes(x = distance, y = PC1, color = POP)) + geom_point() + theme_classic(base_size = 14) + scale_color_manual(values = c("darkred", "navy", "goldenrod4", "gray50"), na.value = "gray70")
```
```{r}
set.seed(12)
# Sample GWAS panels
smry <- summary(df$distance)
## Sample everyone 
dfGWAS <- df %>% sample_n(10000) %>% select("#FID", "IID", "POP")
fwrite(dfGWAS,"../data/SDS/ids/gwas_ids/max.txt", row.names = F, col.names = T, quote = F, sep = "\t")

## 1st quartile
dfGWAS <- df %>% filter(distance <=  smry[2])%>% sample_n(10000) %>% select("#FID", "IID", "POP")
fwrite(dfGWAS,"../data/SDS/ids/gwas_ids/q1.txt", row.names = F, col.names = T, quote = F, sep = "\t")

## median
dfGWAS <- df %>% filter(distance <=  smry[3])%>% sample_n(10000) %>% select("#FID", "IID", "POP")
fwrite(dfGWAS,"../data/SDS/ids/gwas_ids/median.txt", row.names = F, col.names = T, quote = F, sep = "\t")

## 3rd quartile
dfGWAS <- df %>% filter(distance <=  smry[5])%>% sample_n(10000) %>% select("#FID", "IID", "POP")
fwrite(dfGWAS,"../data/SDS/ids/gwas_ids/q3.txt", row.names = F, col.names = T, quote = F, sep = "\t")
```


## Overlap statistics

Load data and process 
```{r}
dfOverlap <- fread("../endpoints/SDS/OverlapStats.txt")
dfOverlap <- dfOverlap %>% mutate(gwas_type  = case_when(gwas == "q1" ~ "q1", 
                                                        gwas == "median" ~ "q2", 
                                                        gwas == "q3" ~ "q3",
                                                        gwas == "max" ~ "q4",
                                                        gwas == "ALL" ~ "ALL",
                                                        gwas == "WBS" ~ "WBS")) %>% 
  mutate(gwas_type2 = case_when(gwas_type %in% c("q1", "q2", "q3", "q4") ~ "distance",
         gwas_type %in% c("ALL", "WBS") ~ "full"))
dfOverlap$gwas_type <- factor(dfOverlap$gwas_type, levels = c("ALL", "WBS", "q1", "q2", "q3", "q4"))
```

Proportion of variance explained (Z) by GWAS type
```{r}
UK <- dfOverlap %>% filter(contrasts == "UK10K") %>% ggplot(aes(x = gwas_type, y = Z, fill = gwas_type2)) + geom_bar(stat = "identity") + xlab("GWAS Type") + ylab("Proportion of variance explained (Z)") + theme_bw(base_size = 14) + scale_fill_manual(values = c("purple4", "lightseagreen"))  + guides(fill="none") + ggtitle("UK10K") + theme(plot.title = element_text(hjust = 0.5)) + ylim(0, 0.0008)

Han <- dfOverlap %>% filter(contrasts == "Han") %>% ggplot(aes(x = gwas_type, y = Z, fill = gwas_type2)) + geom_bar(stat = "identity") + xlab("GWAS Type") + ylab("Proportion of variance explained (Z)") + theme_bw(base_size = 14) + scale_fill_manual(values = c("purple4", "lightseagreen"))  + guides(fill="none") + ggtitle("Han") + theme(plot.title = element_text(hjust = 0.5))+ ylim(0, 0.0008)

# Print plots
p <- ggarrange(UK, Han, nrow = 1)
p
```

Error by GWAS type
```{r}
UK <- dfOverlap %>% filter(contrasts == "UK10K") %>% ggplot(aes(x = gwas_type, y = error, fill = gwas_type2)) + geom_bar(stat = "identity") + xlab("GWAS Type") + ylab("Error") + theme_bw(base_size = 14) + scale_fill_manual(values = c("purple4", "lightseagreen"))  + guides(fill="none") + ggtitle("UK10K") + theme(plot.title = element_text(hjust = 0.5)) + ylim(0,1) + geom_hline(yintercept = 0, color = "red", linetype = 2, linewidth = 1.2)


Han <- dfOverlap %>% filter(contrasts == "Han") %>% ggplot(aes(x = gwas_type, y = error, fill = gwas_type2)) + geom_bar(stat = "identity") + xlab("GWAS Type") + ylab("Error") + theme_bw(base_size = 14) + scale_fill_manual(values = c("purple4", "lightseagreen"))  + guides(fill="none") + ggtitle("Han") + theme(plot.title = element_text(hjust = 0.5)) + ylim(0,1) + geom_hline(yintercept = 0, color = "red", linetype = 2, linewidth = 1.2)

# Print plots
p <- ggarrange(UK, Han, nrow = 1)
p
```

Proportion variance explained by error 
```{r}
UK <- dfOverlap %>% filter(contrasts == "UK10K") %>% ggplot(aes(x = Z, y = error, color = gwas_type)) + geom_point(size = 3) + xlab("Proportion of variance explained (Z)") + ylab("Error") + theme_bw(base_size = 12)  + guides(fill="none") + ggtitle("UK10K") + theme(plot.title = element_text(hjust = 0.5), legend.position = "bottom") + ylim(0,1) + labs(color = "GWAS Type") + geom_hline(yintercept = 0, color = "red", linetype = 2, linewidth = 1.2)

Han <- dfOverlap %>% filter(contrasts == "Han") %>% ggplot(aes(x = Z, y = error, color = gwas_type)) + geom_point(size = 3) + xlab("Proportion of variance explained (Z)") + ylab("Error") + theme_bw(base_size = 12)  + guides(fill="none") + ggtitle("Han") + theme(plot.title = element_text(hjust = 0.5), legend.position = "bottom") + ylim(0,1) + labs(color = "GWAS Type") + geom_hline(yintercept = 0, color = "red", linetype = 2, linewidth = 1.2)

# Print plots
p <- ggarrange(UK, Han, nrow = 1, common.legend = TRUE)
p
```

Compare D by GWAS type  
```{r}
dfOverlap$d_lc <- dfOverlap$D - (2*sqrt(dfOverlap$varD)) 
dfOverlap$d_uc <- dfOverlap$D + (2 *sqrt(dfOverlap$varD))

UK <- dfOverlap %>% filter(contrasts == "UK10K") %>% ggplot(aes(x = gwas_type, y = D, color = gwas_type2)) + geom_point(size = 3) + xlab("GWAS Type") + ylab("D") + theme_bw(base_size = 14) + scale_color_manual(values = c("purple4", "lightseagreen"))  + guides(color="none") + ggtitle("UK10K") + theme(plot.title = element_text(hjust = 0.5)) + geom_hline(yintercept = 0, color = "red", linetype = 2, linewidth = 1.2) + geom_errorbar(aes(ymin = d_lc, ymax = d_uc)) + ylim(0, 2000000000)

Han <- dfOverlap %>% filter(contrasts == "Han") %>% ggplot(aes(x = gwas_type, y = D, color = gwas_type2)) + geom_point(size = 3) + xlab("GWAS Type") + ylab("D") + theme_bw(base_size = 14) + scale_color_manual(values = c("purple4", "lightseagreen"))  + guides(color="none") + ggtitle("Han") + theme(plot.title = element_text(hjust = 0.5)) + geom_hline(yintercept = 0, color = "red", linetype = 2, linewidth = 1.2) + geom_errorbar(aes(ymin = d_lc, ymax = d_uc)) + ylim(0, 2000000000)

# Print plots
p <- ggarrange(UK, Han, nrow = 1, common.legend = TRUE)
p
```

## PC statistics

```{r}
dfPC <- fread("../endpoints/SDS/PCStats.txt")
dfPC <- dfPC %>% mutate(gwas_sample  = case_when(gwas == "q1" ~ "q1", 
                                                        gwas == "median" ~ "q2", 
                                                        gwas == "q3" ~ "q3",
                                                        gwas == "max" ~ "q4",
                                                        gwas == "ALL" ~ "ALL",
                                                        gwas == "WBS" ~ "WBS")) %>% 
  mutate(gwas_sample2 = case_when(gwas_type %in% c("q1", "q2", "q3", "q4") ~ "distance",
         gwas_sample %in% c("ALL", "WBS") ~ "full"))
dfPC$gwas_sample <- factor(dfPC$gwas_sample, levels = c("ALL", "WBS", "q1", "q2", "q3", "q4"))
```

Look at correlation between $PC_i$ and $\hat{F}_{Gr}$
```{r}
# UK
UK <- dfPC %>% filter(gwas_type == "even" & test_type == "odd" & contrasts == "UK10K")%>% ggplot(aes(x = PC, y = B2)) + geom_bar(stat = "identity", color = "darkgreen") + facet_grid(~gwas_sample) + theme_bw(base_size = 12) + ylab(TeX("$B_{i}^2$")) + ggtitle("UK10K") + theme(plot.title = element_text(hjust = 0.5))

# Han
Han <- dfPC %>% filter(gwas_type == "even" & test_type == "odd" & contrasts == "Han")%>% ggplot(aes(x = PC, y = B2)) + geom_bar(stat = "identity", color = "darkgreen") + facet_grid(~gwas_sample) + theme_bw(base_size = 12) + ylab(TeX("$B_{i}^2$")) + ggtitle("Han") + theme(plot.title = element_text(hjust = 0.5))

# Print plots
p <- ggarrange(UK, Han, nrow = 1, common.legend = TRUE)
p
```
Limit to "ALL" 
```{r}
# UK
UK <- dfPC %>% filter(gwas_type == "even" & test_type == "odd" & contrasts == "UK10K" & gwas_sample == "ALL")%>% ggplot(aes(x = PC, y = B2)) + geom_bar(stat = "identity", color = "darkgreen") + facet_grid(~gwas_sample) + theme_bw(base_size = 12) + ylab(TeX("$B_{i}^2$")) + ggtitle("UK10K") + theme(plot.title = element_text(hjust = 0.5)) 


# Han
Han <- dfPC %>% filter(gwas_type == "even" & test_type == "odd" & contrasts == "Han"& gwas_sample == "ALL")%>% ggplot(aes(x = PC, y = B2)) + geom_bar(stat = "identity", color = "darkgreen") + facet_grid(~gwas_sample) + theme_bw(base_size = 12) + ylab(TeX("$B_{i}^2$")) + ggtitle("Han") + theme(plot.title = element_text(hjust = 0.5))

# Print plots
p <- ggarrange(UK, Han, nrow = 1, common.legend = TRUE)
p
```


Limit to "WBS" 
```{r}
# UK
UK <- dfPC %>% filter(gwas_type == "even" & test_type == "odd" & contrasts == "UK10K" & gwas_sample == "WBS")%>% ggplot(aes(x = PC, y = B2)) + geom_bar(stat = "identity", color = "darkgreen") + facet_grid(~gwas_sample) + theme_bw(base_size = 12) + ylab(TeX("$B_{i}^2$")) + ggtitle("UK10K") + theme(plot.title = element_text(hjust = 0.5)) + ylim(0,0.012)

# Han
Han <- dfPC %>% filter(gwas_type == "even" & test_type == "odd" & contrasts == "Han"& gwas_sample == "WBS")%>% ggplot(aes(x = PC, y = B2)) + geom_bar(stat = "identity", color = "darkgreen") + facet_grid(~gwas_sample) + theme_bw(base_size = 12) + ylab(TeX("$B_{i}^2$")) + ggtitle("Han") + theme(plot.title = element_text(hjust = 0.5)) + ylim(0, 0.012)

# Print plots
p <- ggarrange(UK, Han, nrow = 1, common.legend = TRUE)
p
```

Look at correlation between $R^2_{K}$
```{r}
dfPC %>% filter(gwas_type == "even" & test_type == "odd" & contrasts == "UK10K") %>% ggplot(aes(x = PC, y = r2)) + geom_bar(stat = "identity", color = "darkgreen") + facet_grid(~gwas_sample) + theme_bw(base_size = 12) + ylab(TeX("$R_{K}^2$")) + geom_hline(data = dfOverlap, aes(yintercept = gammaF), color = "red", linetype = 2)
```
