---
title: "TestExample"
author: "Jennifer Blanc"
date: "2024-07-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(autodep = TRUE)
library(data.table)
library(tidyverse)
library(ggpubr)
library(latex2exp)
```

# Construct Test and GWAS panels 

## Explore PCA of Whole Biobank

Plot PC1 vs PC2 colored by self-identified ethnic background
```{r, warning=FALSE, message=FALSE}
# Load data and combine
dfPCA <- fread("../endpoints/ukbb/whole_biobank.eigenvec.gz")
dfEB <- fread("../endpoints/ukbb/EthnicBackground_21000.txt")
df <- inner_join(dfPCA, dfEB)

# Use continental labels 
df <- df %>% mutate(continental = case_when(EthnicBackground_21000 %in% c(4, 4001, 4002, 4003) ~ "Africa", EthnicBackground_21000 %in% c(1, 1001, 1002, 1003) ~ "Europe", EthnicBackground_21000 %in% c(2, 2001, 2002, 2003, 2004) ~ "Mixed", EthnicBackground_21000 %in% c(3, 3001, 3002, 3003, 3004) ~ "Asia"))
```

```{r}
# Make PCA plot color based on self identified Ethnic background
ggplot(data = df, aes(x = PC1, y = PC2, color = continental)) + geom_point() + theme_classic(base_size = 14) + scale_color_manual(values = c("darkred", "navy", "goldenrod4", "gray50"), na.value = "gray70")
```
Sample individuals who self-identify as "African" or "European"
```{r}
set.seed(455)
# Sample from each end of PC1
dfEur <- df %>% filter(continental == "Europe") %>% sample_n(10000)
dfAfr <- df %>% filter(continental == "Africa") %>% sample_n(5000)

# Plot data
tmp <- rbind(dfEur, dfAfr)
pl <- ggplot(data = tmp, aes(x = PC1, y = PC2, color = continental)) + geom_point() + theme_classic(base_size = 14) + scale_color_manual(values = c("darkred", "navy", "goldenrod4", "gray50"), na.value = "gray70")
pl
```
Most (though not all) self-identified African individuals fall on one side of PC1 while the self-identified Europeans are on the other side.  

## Sample GWAS panel

Make mixed panels with different amounts of overlap:  

- GWAS panels:  
    - 10,000 Eur  
    - 8,750 Eur, 1,250 Afr  
    - 7,500 Eur, 2,500 Afr  
    - 6,250 Eur, 3,750 Afr  
    - 5,000 Eur, 5,000 Afr  

```{r, eval=FALSE, echo=FALSE}
set.seed(3590)
# Sample GWAS panels
## Complete Overlap
tmpAfr <- dfAfr %>% sample_n(5000) %>% mutate(POP = "Afr")
tmpEur <- dfEur %>% sample_n(5000) %>% mutate(POP = "Eur")
dfGWAS <- rbind(tmpAfr, tmpEur) %>% select("#FID", "IID", "POP")
#fwrite(dfGWAS,"../data/ids/gwas_ids/complete.txt", row.names = F, col.names = T, quote = F, sep = "\t")

## No Overlap
dfGWAS <- dfEur  %>% sample_n(10000) %>% mutate(POP = "Eur")
#fwrite(dfGWAS,"../data/ids/gwas_ids/none.txt", row.names = F, col.names = T, quote = F, sep = "\t")

## Half Overlap
tmpAfr <- dfAfr  %>% sample_n(2500) %>% mutate(POP = "Afr")
tmpEur <- dfEur  %>% sample_n(7500) %>% mutate(POP = "Eur")
dfGWAS <- rbind(tmpAfr, tmpEur) %>% select("#FID", "IID", "POP")
#fwrite(dfGWAS,"../data/ids/gwas_ids/half.txt", row.names = F, col.names = T, quote = F, sep = "\t")

## Quarter
tmpAfr <- dfAfr  %>% sample_n(1250) %>% mutate(POP = "Afr")
tmpEur <- dfEur  %>% sample_n(8750) %>% mutate(POP = "Eur")
dfGWAS <- rbind(tmpAfr, tmpEur) %>% select("#FID", "IID", "POP")
#fwrite(dfGWAS,"../data/ids/gwas_ids/quarter.txt", row.names = F, col.names = T, quote = F, sep = "\t")

## Three Quarters
## GWAS
tmpAfr <- dfAfr  %>% sample_n(3750) %>% mutate(POP = "Afr")
tmpEur <- dfEur  %>% sample_n(6250) %>% mutate(POP = "Eur")
dfGWAS <- rbind(tmpAfr, tmpEur) %>% select("#FID", "IID", "POP")
#fwrite(dfGWAS,"../data/ids/gwas_ids/threequarter.txt", row.names = F, col.names = T, quote = F, sep = "\t")
```

## Explore HGDP1KG data 

```{r}
# Read in data
df <- fread("../data/HGDP1KG/gnomad_meta_updated.tsv", sep = "\t")

# Select relevant columns (for now)
dfFilter <- df %>% select("s","project_meta.project_subpop" , "latitude" ,  "longitude",
                   "population",  "region", "project_meta.project_pop")
colnames(dfFilter)[1] <- "IID"

# Plot data by lat and long
world_map <- map_data("world")
ggplot() + geom_polygon(data = world_map, aes(x = long, y = lat, group = group), fill = "white", color = "black") + geom_point(data = dfFilter, aes(x = longitude, y = latitude, color = project_meta.project_pop), size = 2) +
  theme_minimal(base_size = 14) + xlab("Longitude") + ylab("Latitude") +labs(colour="POP") + theme(legend.position = "bottom")
```

## Sample test panel

Sample 600 inviduals from the "afr" and "nfe" populations
```{r}
set.seed(1212)
dfTest <- dfFilter %>% filter(project_meta.project_pop == "afr" | project_meta.project_pop == "nfe") %>% mutate(POP = case_when(project_meta.project_pop == "afr" ~ "Afr", project_meta.project_pop == "nfe" ~ "Eur")) %>% select("IID", "POP") %>% group_by(POP) %>% sample_n(600) %>% ungroup() %>% mutate(FID = 0) %>% select(FID, IID, POP)
#fwrite(dfTest,"../data/TestExample/ids/test_ids/test.txt", row.names = F, col.names = T, quote = F, sep = "\t")

# Plot data by lat and long
dfPlot <- inner_join(dfFilter, dfTest)
world_map <- map_data("world")
pl <- ggplot() + geom_polygon(data = world_map, aes(x = long, y = lat, group = group), fill = "white", color = "black") + geom_point(data = dfPlot, aes(x = longitude, y = latitude, color = project_meta.project_pop), size = 2) +
  theme_minimal(base_size = 14) + xlab("Longitude") + ylab("Latitude") +labs(colour="POP") + theme(legend.position = "bottom") + scale_color_manual(values = c("darkred", "navy"))
pl
```



# Results 

### $\hat{F}_{Gr}$ for indviduals in the GWAS panel 

```{r}
# Complete Overlap
FGr_complete <- fread("../endpoints/TestExample/completeFGr.txt.gz")
p1 <- ggplot(data = FGr_complete, aes(x = PC1, y = FGr, color = POP)) + geom_point() + theme_classic(base_size = 14) + scale_color_manual(values = c("darkred", "navy")) + ggtitle("Complete Overlap") + theme(plot.title = element_text(hjust = 0.5))

# No Overlap 
FGr_none <- fread("../endpoints/TestExample/noneFGr.txt.gz")
p2 <- ggplot(data = FGr_none, aes(x = PC1, y = FGr, color = POP)) + geom_point() + theme_classic(base_size = 14) + scale_color_manual(values = c("navy")) + ggtitle("No Overlap") + theme(plot.title = element_text(hjust = 0.5))

# Print plots
p <- ggarrange(p1, p2, nrow =1, common.legend = TRUE)
p

ggsave("~/Desktop/tmp.png", p, height = 5, width = 7)
```

### Results 

Load and process overlap data
```{r}
dfOverlap <- fread("../endpoints/TestExample/OverlapStats.txt")
dfOverlap <- dfOverlap %>% mutate(prop_eur  = case_when(gwas == "complete" ~ 0.5, 
                                                        gwas == "none" ~ 1, 
                                                        gwas == "half" ~ 0.75,
                                                        gwas == "quarter" ~ 0.875,
                                                        gwas == "threequarter" ~ 0.625)) 
dfOverlap$L <- factor(dfOverlap$L, levels = c("1000", "3162", "10000", "31623", "100000", "316228", "1000000",  "pruneall"))
dfOverlap <- dfOverlap %>%
  mutate(
    x_numeric = ifelse(L == "pruneall", NA, as.numeric(as.character(L))),
    x_log = log10(x_numeric),
    x_order = ifelse(is.na(x_numeric), Inf, x_numeric)
  ) %>%
  arrange(x_order)
dfOverlap$x <- as.character(dfOverlap$L)
```

Load and process PC data 
```{r}
dfPC <- fread("../endpoints/TestExample/PCStats.txt")
dfPC <- dfPC %>% mutate(prop_eur  = case_when(gwas == "complete" ~ 0.5, 
                                                        gwas == "none" ~ 1, 
                                                        gwas == "half" ~ 0.75,
                                                        gwas == "quarter" ~ 0.875,
                                                        gwas == "threequarter" ~ 0.625))
```


**Overlap vs proportion of variance explained** 
```{r}
tmp1 <- dfOverlap %>% filter(L == "pruneall") %>% select(Z, prop_eur) %>% mutate(type ="FGr")
tmp2 <- dfPC %>% filter(L == "pruneall" & PC == 1 & gwas_type == "all" & test_type == "all") %>% mutate(Z= VarExp) %>% select(Z, prop_eur)%>% mutate(type = "PC 1")  
dfPlot <- rbind(tmp1, tmp2)

pl <- dfPlot %>% ggplot(aes(x = prop_eur, y = Z, color = type)) + geom_point(size = 3) + theme_bw(base_size = 14) + xlab("Proportion EUR in GWAS") + ylab("Proportion of variance explained") + scale_color_manual(values = c("navy", "darkred"), name = "") + theme(legend.position = "bottom")
ggsave("~/Desktop/tmp.png", pl, height = 5, width = 7)
```
As the proportion of overlap between panels decreases, the proportion of variance explained by $\hat{F}_{Gr}$ and PC 1 also decreases.  

**Overlap vs error** 

```{r}
p1 <- dfOverlap %>% filter(L != "pruneall") %>% ggplot(aes(x = as.numeric(as.character(L)), y = error, color = prop_eur))+ geom_point(size = 3) + ylim(0,1.05) + scale_x_log10() + xlab("L")  + ylab("Error") + theme_bw(base_size = 14) +
  theme(legend.position = "none")
p2<- dfOverlap %>% filter(L == "pruneall") %>% ggplot(aes(x = L, y = error, color = prop_eur))+ geom_point(size = 3) + ylim(0,1.05) + xlab("") + ylab("")+ theme_bw(base_size = 14)


pl <- egg::ggarrange(p1, p2, nrow = 1, widths = c(0.75, 0.25))

ggsave("~/Desktop/tmp.png", pl, width = 7, height = 5)
```



```{r}
# Plot all data
p1 <- dfOverlap %>% ggplot(aes(x = prop_eur, y = error, color = L)) + geom_point(size = 3) + theme_bw(base_size = 14) + xlab("Proportion EUR in GWAS") + ylab("Error")  + geom_hline(yintercept = 0, color = "red", linetype = 2, linewidth = 1.2) + ggtitle("All Data") + theme(plot.title = element_text(hjust = 0.5)) 
p1

ggsave("~/Desktop/tmp.png", p1, width = 5, height = 5)
```
```{r}
# Remove no overlap
p2 <- dfOverlap %>% filter(prop_eur %in% c(0.5,0.750, 0.875, 0.625)) %>% ggplot(aes(x = prop_eur, y = error, color = L)) + geom_point(size = 3) + theme_bw(base_size = 14) + xlab("Proportion EUR in GWAS") + ylab("Error") + geom_hline(yintercept = 0, color = "red", linetype = 2, linewidth = 1.2) + ggtitle("Removed No Overlap") + theme(plot.title = element_text(hjust = 0.5))
p2
```
One thing to notice is that while Z decreases gradually with decreased overlap, error goes from 0 to near 1 abruptly with a small amount of overlap. It does appear that error decreases with increased overlap.  


**Proportion variance explained vs error** 
```{r}
# All data
p1 <- dfOverlap %>% ggplot(aes(x = Z, y = error, color = L)) + geom_point(size = 3) + theme_bw(base_size = 12) + ylab("Error") + xlab("Proportion of variance explained") + geom_hline(yintercept = 0, color = "red", linetype = 2, linewidth = 1.2) + ggtitle("All Data") + theme(plot.title = element_text(hjust = 0.5)) + scale_x_log10() + scale_y_log10()

# Remove no overlap
p2 <- dfOverlap %>% filter(prop_eur %in% c(0.5,0.750, 0.875, 0.625)) %>% ggplot(aes(x = Z, y = error, color = L)) + geom_point(size = 3) + theme_bw(base_size = 12) + ylab("Error") + xlab("Proportion of variance explained (Z)") + geom_hline(yintercept = 0, color = "red", linetype = 2, linewidth = 1.2) + ggtitle("Removed No Overlap") + theme(plot.title = element_text(hjust = 0.5)) + scale_y_log10() + scale_x_log10()

pl <- ggpubr::ggarrange(p1, p2, common.legend = TRUE)
pl
```
Not sure there is any new insight from this plot, as Z is related to the overlap. 

**Squared correlation between $PC_i$ between $\hat{F}_{Gr}$**
```{r}
pl <- dfPC %>% filter(gwas_type == "even" & test_type == "odd" & L == "pruneall") %>% ggplot(aes(x = PC, y = B2)) + geom_bar(stat = "identity", fill = "darkgreen", color = "darkgreen") + facet_grid(~prop_eur) + theme_bw(base_size = 12) + ylab(TeX("$B_{i}^2$"))

ggsave("~/Desktop/tmp.png", pl, width = 7, height = 5)
```
As expected, there is a high correlation between PC 1 and $\hat{F}_{Gr}$ in GWAS panels that have overlap. What is a little surprising is that there is still some correlation between PC1 and $\hat{F}_{Gr}$ even when there is no overlap between panels. 


**Cumulative variance explained** 
```{r}
pl <- dfPC %>% filter(gwas_type == "even" & test_type == "odd" & L == "pruneall") %>% ggplot(aes(x = PC, y = r2)) + geom_bar(stat = "identity", fill = "darkgreen", color = "darkgreen") + facet_grid(~prop_eur) + theme_bw(base_size = 12) + ylab(TeX("$R_{K}^2$")) + geom_hline( aes(yintercept = 1-Error), color = "red", linetype = 2, linewidth = 1.5)

ggsave("~/Desktop/tmp.png", pl, width = 7, height = 5)
```

```{r, echo=FALSE, eval=FALSE}
df <- fread("~/Downloads/fourier_ls-all_parsed.bed")
dfBB <- df %>% group_by(chr)  %>%
  mutate(block_group = (block_number - 1) %/% 3) %>%
  group_by(chr, block_group) %>%
  summarize(
    start = min(start),
    stop = max(stop),
    block_number = paste(unique(block_number), collapse = "-"),
    .groups = 'drop'
  )
dfBB$block_number <- seq(1, nrow(dfBB))
dfBB <- dfBB %>% select(chr, start, stop, block_number)

fwrite(dfBB,"~/polygenic_selection_stratification/data/LD_Blocks/big_blocks.bed", row.names = F, col.names = T, quote = F, sep = "\t")
```




